---
runtime: shiny
title: "Monte carlo simulation"
output: flexdashboard::flex_dashboard
---


```{r}
knitr::opts_chunk$set(echo=F, message=TRUE, warning=TRUE)
library(ggplot2)
library(DT)
```

```{r}
per_step = 10

rad=1
min=-1
max=min+2*rad

angle = seq(-pi, pi, length = 50)
circle = data.frame(x = sin(angle), y = cos(angle))
```

Sidebar {.sidebar}
=====================================

The aim of this exercise is to simulate the value of $\pi$ (`r pi`).

First, we draw a circle and a square.<br>

Second, we start the simulation.

1. Randomly draw x,y coordinates in square.
2. Check if coordinates falls within the circle using formula
$x^2 + y^2 < R^2$, where $R$ is the radius of the circle. 
3. Tally number of points in circle/ number of points in square. Values should converge on $\frac{\pi}{4}$.

```{r}
actionButton("add", "Step", style="background-color: #337ab7")
actionButton("reset", "Reset")
numericInput("step", "Step size (power of 10)", 1, min = 1, max = 3)
```

- See tab 2 for data generated.
- See tab 3 tab for why this works.


```{r}
values = reactiveValues()
values$counter = 0
values$data = data.frame(x=numeric(), y=numeric(), inside=integer())
values$step_size = per_step

observeEvent(input$add, { 
    values$step_size = per_step^input$step 
    values$counter = values$counter + values$step_size
    x = runif(values$step_size, min=min, max=max)
    y = runif(values$step_size, min=min, max=max)
    inside = ifelse(x^2 + y^2 < max^2, 1, 0) 
    values$data = rbind(values$data, data.frame(x=x, y=y, inside=inside))
})
observeEvent(input$reset, {
    values$counter = 0
    values$data = data.frame(x=numeric(), y=numeric(), inside=integer())
})

output$data = renderDataTable({values$data})
```

Plot
=====================================  

```{r}

fluidPage(
    fluidRow(
        column(6, 
            renderPlot({
                ggplot(values$data, aes(x=x, y=y, color=factor(inside))) + 
                    geom_polygon(aes(x, y), data = circle, inherit.aes = F) + 
                    geom_point() +
                    theme_minimal() +
                    scale_x_continuous(limits=c(min,max)) +
                    scale_y_continuous(limits=c(min,max)) +
                    guides(color=FALSE) 
            }, height=300, width=300)),
        column(6,
            fluidPage(
                renderText({ 
                    paste("Counter: ", values$counter, " points added")
                }),
                renderText({ 
                    paste("Currently: ", values$step_size, " points added at each step")
                }),
                renderText({
                    if(dim(values$data)[1] >= 1) {
                        pi_sim = 4*sum(values$data$inside==1)/length(values$data$inside)
                        paste("Simulated pi = ", round(pi_sim, 7))
                    }
                })
                )
)))
```

Data
=====================================  
```{r}
dataTableOutput("data")
```

Why this works
=====================================  

This is because <br>

$A_s = 2R * 2R = 4R^2$ <br>
$A_c = \pi * R^2$ <br>
$\frac{A_c}{A_s} = \frac{\pi * R^2}{4R^2} = \frac{\pi}{4}$ <br>

Thus, $\pi = 4 * \frac{number of points in circle}{number of points in square}$
