---
runtime: shiny
title: "COVID 19 Singapore"
output: 
  html_document:
    toc: true
    theme: flatly
---

```{r}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```

```{r echo=TRUE}
# Libraries
library(rvest)
library(ggplot2)
library(tidyr)
library(data.table)
library(gridExtra)
source('main.r')
source('util.r')

# Resources
# https://staff.math.su.se/hoehle/blog/2020/03/16/flatteningthecurve.html
# https://blog.ephorie.de/covid-19-the-case-of-germany 
# https://rviews.rstudio.com/2020/03/19/simulating-covid-19-interventions-with-r/

# Params
theme = theme_minimal()
```

## Get data

```{r echo=TRUE}
url = "https://co.vid19.sg/cases/search"
df = tryCatch({ get_data(url) })
```

Data is obtained from `r url`. 

## Clean data

```{r echo=TRUE}
if(exists("df")) {
  df = clean_data(df)
} else {
  df = get_alternative_data()
}  
str(df)
```

### Notes about data

1. symptomatic < confirmed < recovered date
covid19-coronavirus-singapore-discharged-patients-moh-12477696). 
2. `days.to.recover.symptomatic` = recovered - symptomatic
3. `days.to.recover.resources` = recovered - confirmed

## What are the figures

### Coronavirus through time

#### Cases detected through time
```{r}
min_date = min(df$confirmed.at.date, na.rm=T)
max_date = max(df$confirmed.at.date, na.rm=T)
date_second_wave = as.Date("2020-02-25")
```

The first case detected in Singapore on `r format(min_date, "%d %B %Y")`. 

```{r}
style = "display: inline-block;vertical-align:top;padding-right: 20px"
bootstrapPage(
  div(style=style, dateRangeInput("range_date", "Date range", 
                                  start = min_date, end = max_date,
                                  min = min_date, max = max_date)),
  div(style=style, numericInput("binwidth_date", "Days bin width", value = 1, min=1, max=14)))

renderPlot({
    # input = list(range_date = c(as.Date(min_date), as.Date(max_date)), binwidth=1) # test
    min_date = input$range_date[1]
    max_date = input$range_date[2]
    data = df[df$confirmed.at.date >= min_date & df$confirmed.at.date <= max_date,]

    ggplot(data=data, aes(x=confirmed.at.date)) +
      geom_histogram(stat="bin", binwidth=input$binwidth_date, fill="steelblue") + 
      xlab("") + ylab("") + 
      geom_vline(xintercept=date_second_wave, size=2, col="red") +
      scale_x_date(date_breaks = "weeks", date_labels = "%Y EW %U") +
      theme
}, height=280)

renderPlot({
    min_date = input$range_date[1]
    max_date = input$range_date[2]
    data = df[df$confirmed.at.date >= min_date & df$confirmed.at.date <= max_date,]

    type_label = c("imported case", "local transmission")
    names(type_label) = c("Imported", "Local")

    ggplot(data=data, aes(x=confirmed.at.date, fill=infection.source)) +
      geom_histogram(stat="bin", binwidth=input$binwidth_date) + 
      facet_wrap(~ infection.source, ncol=1, labeller = camelcase_labeller) + 
      geom_vline(xintercept=date_second_wave, size=2, col="red") +
      scale_fill_discrete(name="Case type", breaks=type_label, labels=names(type_label)) + 
      xlab("") + ylab("Number of confirmed cases detected\n") +
      scale_x_date(date_breaks = "weeks", date_labels = "%Y EW %U") +
      theme
})
```

From here, we can see that there are two waves of cases, the first from EW 03 to EW 07 and second from EW 08 onwards. These are associated with the Chinese cases mainly and when there was a global spread subsequently with Europe as the epicenter.


```{r}
# data_1st_wave = data.table(df[df$confirmed.at.date < date_second_wave &
#                               df$infection.source == "imported case",])
# data_1st_wave = data_1st_wave[, list(N=length(case)),by="country.of.origin"]

# data_2nd_wave = data.table(df[df$confirmed.at.date >= date_second_wave &
#                               df$infection.source == "imported case",])
# data_2nd_wave = data_2nd_wave[, list(N=length(case)),by="country.of.origin"]

# grid.arrange(
#   ggplot(data_1st_wave, aes(x=camelcase_labeller("", country.of.origin), y=N)) +
#   geom_bar(stat = "identity") + coord_flip() + theme_minimal() +
#   xlab("") + ylab("") + ggtitle("First wave"),
#   ggplot(data_2nd_wave, aes(x=camelcase_labeller("", country.of.origin), y=N)) +
#   geom_bar(stat = "identity") + coord_flip() + theme_minimal() +
#   xlab("") + ylab("Number of imported cases detected") + ggtitle("Second wave")
# )
```


#### Cases through time

Case load is an important metric that is correlated also with hospital load. Here, it's the sum of people who are not recovered. 

Measuring hospital load is key to ensure that hospitals are not overloaded.  We assume that the number of those who require hospitalisation is 20% of the case load, since it is a conservative estimate that 20% of the patients require hospitalisation and/or ICU care. From [24 Mar 2020](https://go.gov.sg/measures24mar), those who still test positive but display no symptoms will no longer be hospitalised. 

There are 9 public acute hospitals in Singapore, with an average capacity of 1,200 beds. They are currently operating at approximately [75% load](https://www.moh.gov.sg/resources-statistics/healthcare-institution-statistics/beds-occupancy-rate-(bor)). We assume that hospitals can operate at a 90% bed load, to ensure that hospitals are not overtaxed and there is sufficient isolation between covid patients (more than usual required). Then there would be `r # max_load` beds available. 

```{r}

# N cases still requiring treatment

# date_no_hosp = as.Date("2020-03-24")
# df_cases = df[df$status != "Deceased", c("case", "symptomatic.at.date", "recovered.at.date", "confirmed.at.date")]
# df_cases[is.na(df_cases$recovered.at.date),]$recovered.at.date = Sys.Date()
# df_cases[is.na(df_cases$symptomatic.at.date),]$symptomatic.at.date = 
#   df_cases[is.na(df_cases$symptomatic.at.date),]$confirmed.at.date # use confirmed.at if no symptomatic

# df_cases$day = mapply(function(min, max) { seq(min, max, by="day") }, 
#                        min=df_cases$symptomatic.at.date, max=df_cases$recovered.at.date)
# df_cases_day = data.table(unnest(df_cases, cols=c(day, case))); rm(df_cases)
# df_cases_day = df_cases_day[, list(hospital_load=length(case)), by="day"]
# df_cases_day$hospital_load = as.numeric(df_cases_day$hospital_load)
# df_cases_day[day>=date_no_hosp,]$hospital_load =  df_cases_day[day>=date_no_hosp,]$hospital_load*0.2


# max_load = 9*1200*0.15
# load_prop = 0.2

# renderPlot({
#   ggplot(data=df_cases_day, aes(x=day, y=hospital_load)) +
#     geom_bar(stat="identity", fill="steelblue") + 
#     xlab("") + ylab("Number of hospitalised peersons\n") + 
#     scale_x_date(date_breaks = "weeks" , date_labels = "%Y EW %U") +
#     theme
# }, height=280)

# renderPlot({
#   ggplot(data=df_cases_day, aes(x=day, y=round(hospital_load/max_load*100, 2))) +
#     geom_bar(stat="identity", fill="steelblue") + 
#     geom_vline(xintercept=date_no_hosp-1, size=2, col="red") +
#     xlab("") + ylab("% of (estimated) maximum patient load\n") + 
#     scale_x_date(date_breaks = "weeks" , date_labels = "%Y EW %U") +
#     theme
# }, height=280)
```

### Recovery time

Recovered date is based on the discharged date when there are [consecutive negative PCR tests](https://www.channelnewsasia.com/news/singapore/.

#### Days to recovery histogram

```{r}
# is.recovered = table(!is.na(df[, c("days.to.recover.symptomatic")]))
```
- Number of recovered persons: `r # is.recovered[2]`
- Proportion of recovered persons: `r # round(prop.table(is.recovered)[2]*100, 2)`%

- Median time to recovery (from confirmed): `r # as.numeric(median(df[, c("days.to.recover.symptomatic")], na.rm=T))` days
- Median time to recovery (from confirmed): `r # round(as.numeric(mean(df[, c("days.to.recover.symptomatic")], na.rm=T)), 2)` days

```{r fig.height=3, fig.width=5}
# binwidth = 5
# min = as.numeric(min(df[, c("days.to.recover.symptomatic")], na.rm=T))
# max = as.numeric(max(df[, c("days.to.recover.symptomatic")], na.rm=T))
# hist_max = max(table(cut(as.numeric(df[, c("days.to.recover.symptomatic")]), 
#                           breaks=seq(min, max, by=binwidth), right = FALSE)))
# ggplot(df, aes(x=days.to.recover.symptomatic)) + geom_histogram(binwidth=binwidth, na.rm=T) + 
#   xlab("Days to recover") + ylab("Number of people") +
#   scale_y_continuous(breaks=seq(0, hist_max, ifelse(hist_max<30, 1, 5))) +
#   scale_x_continuous(breaks=seq(0, max, ifelse(max<30, 1, 5))) + 
#   theme
```

#### Association between age and days to recover

Here we attempt to investigate if there is an association with number of days to recover and age, as [fatalities are higher for elderly above 65](https://www.businessinsider.sg/most-us-coronavirus-deaths-ages-65-older-cdc-report-2020-3).

```{r}
# selectInput("age_width", label = "Age bin width", choices = c(5, 10, 20), selected = 5)
# renderPlot({
#   # input = list(age_width=10)
#   agewidth = as.numeric(input$age_width)
#   min = as.numeric(min(df$age, na.rm=T))
#   max = as.numeric(max(df$age, na.rm=T))
#   data = data.frame(cat=cut(as.numeric(df$age), breaks=seq(0, max, by=agewidth), right = FALSE),
#                     age=df$age, 
#                     days=df[, c("days.to.recover.symptomatic")])
#   data = data[!is.na(data$cat),]

#   ggplot(data, aes(x=cat, y=days)) + geom_boxplot() +
#     geom_jitter(height = 0, width = 0.1) +
#     xlab("Age category") + ylab("Days to recover") + theme
# })
```

```{r fig.height=4, fig.width=12}
# ggplot(df, aes(x=as.numeric(age), y=as.numeric(days.to.recover.symptomatic))) + 
#   geom_smooth(method='lm', formula= y~x) +
#   geom_point() + 
#   xlab("Age") + ylab("Days to recover") +
#   theme

# df_corr = df[!is.na(df[, c("days.to.recover.symptomatic")]),]
# rsq = cor(as.numeric(df_corr$age), as.numeric(df_corr$days.to.recover.symptomatic))^2
```
Age does not appear to be associated with the number of days to recover ($R^2$ = `r #round(rsq*100, 2)`%). 

### Interventions in Singapore

#### Key measures

- 2019-03-28: defer non-essential trips to malls
- 2019-03-27: [home based learning](http://go.gov.sg/homebasedlearning) once a week for primary, secondary schools, and junior college; staggered Wednesday, Thursday, Friday respectively 
- 

Source: https://www.moh.gov.sg/news-highlights/